Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-01-07 analysis/paper.tex
Add introduction and expand discussion

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  review]{elsarticle}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources over in
terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3
ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

    \begin{itemize}
    \tightlist
    \item[$\square$]
      General trends
    \item[$\square$]
      Case studies
    \item[$\boxtimes$]
      Comparison with archaeobot
    \end{itemize}
  \item[$\square$]
    Conclusion
  \end{itemize}
\item[$\square$]
  Figures

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    fig-region
  \item[$\boxtimes$]
    fig-occ
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \end{itemize}
\item[$\square$]
  Outstanding TODOs
\item[$\square$]
  Add missing references
\item[$\square$]
  Copyedit
\item[$\square$]
  Appendix with all hindcast predictions (HTML?)
\item[$\square$]
  Clean up package code
\item[$\square$]
  Final proofread
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  The first farming societies had an ecological context
\item
  Subsistence is understood (largely) through archaeobotany and
  zooarchaeology; ecological context from environmental archaeology,
  palynology, palaeoclimate records, etc.

  \begin{itemize}
  \tightlist
  \item
    But these have a variety of biases (human selection, taphonomy,
    etc.)
  \item
    And at the end of the day only represent specific places --
    interpolation to the entire region is not straightforward
  \item
    Overlap makes it difficult to see where human choices depart from
    the environmental background \citep[cf.][]{MartinEtAl2016}
  \item
    Or to fully contextualise subsistence strategies and shifts in
    strategies in response to environmental shifts
    \citep[e.g.][]{YaworskyEtAl2023}.
  \end{itemize}
\item
  Here we present an alternative approach using ENM

  \begin{itemize}
  \tightlist
  \item
    Whole region, at multiple climate snapshots
  \item
    Independent of archaeobot. and pal. clim. data, so can verify and
    compare
  \end{itemize}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    Distinguish environmental from potentially anthropogenic change
    \citep{MartinEtAl2017, MartinEtAl2025}
  \item
    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

@@

\citet{FranklinEtAl2015} review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. In an
early application to West Asia, \citet{ConollyEtAl2012} used the
occurrence of wild and domestic \emph{Bos} remains at prehistoric
archaeological sites to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales
\citep[e.g.][]{deAndresHerreroEtAl2018, YaworksyEtAl2023} and, in a West
Asian context, of foraged plant resources in the landscape around the
Neolithic site of XX \citep{CollinsEtAl2018}. Modelling the spread of
crops has been another significant archaeological application
\citep{CremaEtAl}.

@@

\section{Methods and materials}\label{methods-and-materials}

@@

@@

\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrr}
\toprule
 & \multicolumn{2}{c}{Occurrences} \\ 
\cmidrule(lr){2-3}
Taxon & Neolithic & Present \\ 
\midrule\addlinespace[2.5pt]
\emph{Triticum turgidum dicoccum}

incl. \emph{Triticum aestivum}, \emph{Triticum dicoccoides}, \emph{Triticum dicoccum} & 64 (44\%) & 200 \\ 
\emph{Hordeum spontaneum}

incl. \emph{Hordeum vulgare} & 62 (43\%) & 6912 \\ 
\emph{Triticum monococcum aegilopoides}

incl. \emph{Triticum boeoticum}, \emph{Triticum monococcum} & 44 (30\%) & 3213 \\ 
\emph{Bolboschoenus maritimus}

incl. \emph{Scirpus maritimus} & 33 (23\%) & 364 \\ 
\emph{Vicia ervilia} & 33 (23\%) & 858 \\ 
\emph{Buglossoides tenuiflora} & 31 (21\%) & 159 \\ 
\emph{Arnebia decumbens} & 25 (17\%) & 240 \\ 
\emph{Buglossoides arvensis} & 25 (17\%) & 251 \\ 
\emph{Medicago radiata} & 21 (14\%) & 883 \\ 
\emph{Androsace maxima} & 20 (14\%) & 130 \\ 
\emph{Vicia orientalis}

incl. \emph{Lens culinaris} & 17 (12\%) & 441 \\ 
\emph{Medicago astroites}

incl. \emph{Trigonella astroites} & 16 (11\%) & 130 \\ 
\emph{Arnebia linearifolia} & 14 (10\%) & 158 \\ 
\emph{Linum bienne}

incl. \emph{Linum usitatissimum} & 11 (8\%) & 274 \\ 
\emph{Gypsophila vaccaria}

incl. \emph{Vaccaria pyramidata} & 11 (8\%) & 326 \\ 
\emph{Carex divisa} & 10 (7\%) & 330 \\ 
\emph{Ficus carica} & 10 (7\%) & 4265 \\ 
\emph{Lathyrus oleraceus}

incl. \emph{Pisum elatius}, \emph{Pisum sativum} & 8 (6\%) & 2269 \\ 
\emph{Vicia faba} & 8 (6\%) & 2578 \\ 
\emph{Aizoanthemopsis hispanica}

incl. \emph{Aizoon hispanicum} & 7 (5\%) & 303 \\ 
\emph{Bolboschoenus glaucus} & 7 (5\%) & 27\textsuperscript{\textit{*}} \\ 
\emph{Pistacia atlantica} & 7 (5\%) & 928 \\ 
\emph{Polygonum arenarium arenarium}

incl. \emph{Polygonum venantianum} & 7 (5\%) & 5\textsuperscript{\textit{*}} \\ 
\emph{Prosopis farcta} & 7 (5\%) & 1205 \\ 
\emph{Rumex pulcher} & 7 (5\%) & 794 \\ 
\emph{Ammi majus} & 6 (4\%) & 461 \\ 
\emph{Cicer reticulatum}

incl. \emph{Cicer arietinum} & 6 (4\%) & 193 \\ 
\emph{Hordeum bulbosum} & 6 (4\%) & 2036 \\ 
\emph{Polygonum corrigioloides} & 6 (4\%) & 4\textsuperscript{\textit{*}} \\ 
\emph{Salsola kali} & 6 (4\%) & 369 \\ 
\emph{Taeniatherum caput-medusae} & 6 (4\%) & 235 \\ 
\emph{Capparis spinosa} & 5 (3\%) & 4031 \\ 
\emph{Chenopodium album} & 5 (3\%) & 357 \\ 
\emph{Lolium temulentum} & 5 (3\%) & 148 \\ 
\emph{Poa bulbosa} & 5 (3\%) & 1401 \\ 
\emph{Aegilops crassa} & 4 (3\%) & 170 \\ 
\emph{Atriplex prostrata} & 4 (3\%) & 112 \\ 
\emph{Brachypodium distachyon} & 4 (3\%) & 1581 \\ 
\emph{Fumaria densiflora} & 4 (3\%) & 194 \\ 
\emph{Gypsophila pilosa} & 4 (3\%) & 79 \\ 
\emph{Hordeum murinum} & 4 (3\%) & 1342 \\ 
\emph{Lathyrus aphaca} & 4 (3\%) & 2085 \\ 
\emph{Lathyrus sativus} & 4 (3\%) & 297 \\ 
\emph{Secale cereale} & 4 (3\%) & 1153 \\ 
\emph{Triticum durum} & 4 (3\%) & 219 \\ 
\emph{Vitis sylvestris} & 4 (3\%) & 3\textsuperscript{\textit{*}} \\ 
\emph{Adonis flammea} & 3 (2\%) & 79 \\ 
\emph{Avena sterilis} & 3 (2\%) & 50001\textsuperscript{\textit{*}} \\ 
\emph{Bassia arabica} & 3 (2\%) & 230 \\ 
\emph{Bromus sterilis} & 3 (2\%) & 385 \\ 
\emph{Cephalaria syriaca} & 3 (2\%) & 144 \\ 
\emph{Glaucium aleppicum} & 3 (2\%) & 22\textsuperscript{\textit{*}} \\ 
\emph{Helianthemum ledifolium} & 3 (2\%) & 668 \\ 
\emph{Lolium rigidum} & 3 (2\%) & 1464 \\ 
\emph{Phalaris paradoxa} & 3 (2\%) & 370 \\ 
\emph{Quercus ithaburensis} & 3 (2\%) & 1431 \\ 
\emph{Triticum aestivum compactum} & 3 (2\%) & 209 \\ 
\emph{Verbena officinalis} & 3 (2\%) & 1046 \\ 
\emph{Vicia narbonensis} & 3 (2\%) & 1529 \\ 
\emph{Triticum urartu} & — & 2033 \\ 
\emph{Aegilops speltoides} & — & 1429 \\ 
\emph{Aegilops tauschii} & — & 1395 \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{*}}Excluded from modelling due to sample size\\
\end{minipage}

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia for the \texttt{r\ nrow(flora)} modelled taxa
(N=\texttt{r\ nrow(unnest(flora,\ occ))})}

@@

@@

\subsection{Model assessment}\label{model-assessment}

\begin{verbatim}
Warning: Removed 51 rows containing missing values or values outside the scale range
(`geom_text_repel()`).
\end{verbatim}

@@

\caption{\label{fig-model-sensitivity-vs-n_occ}Number of taxon
occurrences and model sensitivity on test dataset}

@@

Many taxa that occur (or are predicted to occur) across the `hilly
flanks' today---including most crop progenitors---are reconstructed to
have had a significantly more restricted distribution in the terminal
Pleistocene/Early Holocene. These include \emph{Ficus carica} (fig);
\emph{Hordeum} spp. (wild barleys); \emph{Lathyrus aphaca} and \emph{L.
sativus} (both marginally edible legumes); \emph{Triticum aestivum
compactum} (in the N. Levant), \emph{T. monococcum aegilopoides},
\emph{T. durum}, and \emph{Triticum urartu} (but not the other wheat
progenitor, \emph{T. turgidum dicoccum} -- see below); \emph{Aegilops
speltoides}, but not \emph{Aegilops tauschii} (goatgrasses); and
\emph{Vicia} spp. (vetches), including \emph{Vicia faba} (broad beans).
Central Anatolia, Northern Mesopotamia, and the Zagros Mountains in
particular disappear from the predicted ranges of these species, leaving
the Levant and to a lesser extent Western Anatolia Cyprus as refugia.

Our results for the Levant are consistent with the current understanding
of this region as developing early intensive foraging economies {[}the
Natufian culture, @{]} and as a centre of origin of agriculture {[}@{]}.
It should also be noted that the selection of flora for this study
(Table~\ref{tbl-occ-count}) is biased towards Levantine fauna due to the
history of archaeological research in the region, as is the occurrence
dataset used to train the model Figure~\ref{fig-occ-map}. Within the
Levant, retreat from the Badia.

Loss of the Northern Mesopotamia--Southeastern Anatolia region from the
predicted ranges of crop progenitors is interesting in light of the
`golden triangle' hypothesis \citep{KozlowskiAurenche2005}, which puts
this region at the centre of the development of agriculture. Multiple
lines of archaeological evidence have emerged that point away from this
hypothesis and towards a more geographically diverse origin
\citep{Asouti2006, cites}, but it remains the area with some of the
earliest clear evidence of domestication. Comparative genetics also
points to the Northern Mesopotamia region as the centre of diversity of
many crops \citep{cite}. But since these studies are based on modern
genomes, if the wild range of these plants has, as our modelling
suggests, shifted since the Early Holocene, the apparent centre of
diversity may have shifted with them. Our reconstructions are consistent
with the late arrival of intensive plant-based foraging economies in
this region (cf.~the Natufian of the Levant) and more broadly there need
not be a link between the core \emph{wild} range of a plant and the core
area of its domestication. A scenario in which cultivation emerged at
the edges of the ranges of valuable plant resources---as a means of
extend their natural niche---is also plausible.

@@

Cyprus and W. Anatolia not considered part of the core zone -- maybe
they should be?

Exceptions to the trend of range reduction include \emph{Cicer
reticulatum} (wild chickpea), which has a relatively stable range
centered on Northern Mesopotamia; and \emph{Triticum turgidum dicoccum}
(wild emmer wheat), which is predicted to have two very restricted
ranges centered around the Black Sea Coast of Anatolia and the Palmyra
basin. In the latter case, neither of these areas are part of the
predicted modern distribution of wild emmer (centered around the
Caucasus and Northern Mesopotamia), but it would be consistent with
archaeological evidence for early cultivation at Tell Abu Hureyra and
other sites in the Upper Euphrates. On the other hand, the model for
\emph{Triticum turgidum dicoccum} has the lowest predictive performance
of all our fitted models.

@@

Several crop progenitors reduced in range before cultivation? Reduction
in range from BA to EH very evident in Hordeum, Pistacia atlantica.

@@

\section{Conclusion}\label{conclusion}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint]{elsarticle}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in the
terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

    \begin{itemize}
    \tightlist
    \item[$\boxtimes$]
      General trends
    \item[$\square$]
      Case studies
    \item[$\boxtimes$]
      Comparison with archaeobot
    \end{itemize}
  \item[$\square$]
    Conclusion
  \end{itemize}
\item[$\square$]
  Figures

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    fig-region
  \item[$\boxtimes$]
    fig-occ
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-model-metrics
  \end{itemize}
\item[$\square$]
  Outstanding TODOs
\item[$\square$]
  Add missing references
\item[$\square$]
  Copyedit
\item[$\square$]
  Appendix with all hindcast predictions (HTML?)
\item[$\square$]
  Clean up package code
\item[$\square$]
  Final proofread
\end{itemize}

@@

The Pleistocene--Holocene transition in West Asia marked a turning point
in global environmental history, as humans brought the first plants
under cultivation and began modifying surrounding ecosystems to support
their own subsistence. West Asia is part of the native range of a
remarkable number of domesticable plant species, including wild
relatives of wheat, barley, peas, lentils, and other crops of global
importance \citep{cite}. Even before the end of the Ice Age, these
species supported uniquely dense and complex Late Epipalaeolithic
(15--11.7 ka) societies based on intensive foraging \citep{cite} and
eventually pre-domestication cultivation \citep{cite}. As they were
brought under domestication in the Pre-Pottery Neolithic period
(11.7--8.5 ka), the world's first agriculture was shaped by the
ecosystems from which it emerged and was embedded in.

Decades of research in archaeobotany and zooarchaeology have
reconstructed the subsistence economies of Late Epipalaeolithic and
Neolithic sites in great detail \citep{cites}. Together with studies of
charcoal \citep{cite}, pollen \citep{cite}, soil \citep{cite}, and a
variety of palaeoclimate archives \citep{JonesEtAl2019}, they also tell
us much about the environments surrounding these settlements. However,
each of these sources of evidence is subject to the wide variety of
taphonomic and recovery biases that are inherent in any direct record of
the past. They are also, by definition, records of the (human)
environment at particular times and places. Interpolating these
snapshots to give a holistic picture of the regional ecologies is not
straightforward -- to date, it has tended to rely on non-explicit,
inductive modelling. The majority are also filtered through human
action, producing a mixed single that makes it difficult to disentangle
anthropic effects from the background of environmental change in this
period of rapid climatic alteration.

In this paper we present a complementary, deductive approach based on
ecological niche modelling. Rather than inferring environmental
conditions from preserved physical evidence, we predict the ranges of
individual species relevant to human subsistence based on a model of
their current environmental niche and simulations of past palaeoclimate.
Though hypothetical, this gives us an independent line of evidence on
past ecologies that is independent of the environmental archaeological
and palaeoclimatic records. Our computational approach is also readily
scaled up, allowing us to model spatially-explicit palaeodistributions
for a large number of species, for the whole region, under multiple past
climatologies.

@@

  \begin{itemize}
  \tightlist
  \item
    Distinguish environmental from potentially anthropogenic change
    \citep{MartinEtAl2016, MartinEtAl2025}
  \item
    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

@@

\citet{FranklinEtAl2015} review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. In an
early application to West Asia, \citet{ConollyEtAl2012} used the
occurrence of wild and domestic \emph{Bos} remains at prehistoric
archaeological sites to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales
\citep[e.g.][]{deAndresHerreroEtAl2018, YaworskyEtAl2023} and, in a West
Asian context, of foraged plant resources in the landscape around the
Neolithic site of XX \citep{CollinsEtAl2018}. Modelling the spread of
crops has been another significant archaeological application
\citep{CremaEtAl}.

@@

\section{Data and model}\label{data-and-model}

@@

\begingroup
\setlength\LTleft{0\linewidth}
\setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont
\setlength{\LTpost}{0mm}

\begin{longtable}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 0.50\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 7.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 7.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 15.00pt -2\tabcolsep-1.5\arrayrulewidth}}

@@

\tabularnewline

\toprule
 & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 30.00pt -2\tabcolsep-1.5\arrayrulewidth}}{Occurrences} \\ 
\cmidrule(lr){2-4}
Taxon & Neolithic & (\%) & Present \\ 
\midrule\addlinespace[2.5pt]
\emph{Triticum turgidum dicoccum}

incl. \emph{Triticum aestivum}, \emph{Triticum dicoccoides}, \emph{Triticum dicoccum} & 64 & 44\% & 200 \\ 
\emph{Hordeum spontaneum}

incl. \emph{Hordeum vulgare} & 62 & 43\% & 6912 \\ 
\emph{Triticum monococcum aegilopoides}

incl. \emph{Triticum boeoticum}, \emph{Triticum monococcum} & 44 & 30\% & 3213 \\ 
\emph{Bolboschoenus maritimus}

incl. \emph{Scirpus maritimus} & 33 & 23\% & 364 \\ 
\emph{Vicia ervilia} & 33 & 23\% & 858 \\ 
\emph{Buglossoides tenuiflora} & 31 & 21\% & 159 \\ 
\emph{Arnebia decumbens} & 25 & 17\% & 240 \\ 
\emph{Buglossoides arvensis} & 25 & 17\% & 251 \\ 
\emph{Medicago radiata} & 21 & 14\% & 883 \\ 
\emph{Androsace maxima} & 20 & 14\% & 130 \\ 
\emph{Vicia orientalis}

incl. \emph{Lens culinaris} & 17 & 12\% & 441 \\ 
\emph{Medicago astroites}

incl. \emph{Trigonella astroites} & 16 & 11\% & 130 \\ 
\emph{Arnebia linearifolia} & 14 & 10\% & 158 \\ 
\emph{Linum bienne}

incl. \emph{Linum usitatissimum} & 11 & 8\% & 274 \\ 
\emph{Gypsophila vaccaria}

incl. \emph{Vaccaria pyramidata} & 11 & 8\% & 326 \\ 
\emph{Carex divisa} & 10 & 7\% & 330 \\ 
\emph{Ficus carica} & 10 & 7\% & 4265 \\ 
\emph{Lathyrus oleraceus}

incl. \emph{Pisum elatius}, \emph{Pisum sativum} & 8 & 6\% & 2269 \\ 
\emph{Vicia faba} & 8 & 6\% & 2578 \\ 
\emph{Aizoanthemopsis hispanica}

incl. \emph{Aizoon hispanicum} & 7 & 5\% & 303 \\ 
\emph{Bolboschoenus glaucus} & 7 & 5\% & 27\textsuperscript{\textit{*}} \\ 
\emph{Pistacia atlantica} & 7 & 5\% & 928 \\ 
\emph{Polygonum arenarium arenarium}

incl. \emph{Polygonum venantianum} & 7 & 5\% & 5\textsuperscript{\textit{*}} \\ 
\emph{Prosopis farcta} & 7 & 5\% & 1205 \\ 
\emph{Rumex pulcher} & 7 & 5\% & 794 \\ 
\emph{Ammi majus} & 6 & 4\% & 461 \\ 
\emph{Cicer reticulatum}

incl. \emph{Cicer arietinum} & 6 & 4\% & 193 \\ 
\emph{Hordeum bulbosum} & 6 & 4\% & 2036 \\ 
\emph{Polygonum corrigioloides} & 6 & 4\% & 4\textsuperscript{\textit{*}} \\ 
\emph{Salsola kali} & 6 & 4\% & 369 \\ 
\emph{Taeniatherum caput-medusae} & 6 & 4\% & 235 \\ 
\emph{Capparis spinosa} & 5 & 3\% & 4031 \\ 
\emph{Chenopodium album} & 5 & 3\% & 357 \\ 
\emph{Lolium temulentum} & 5 & 3\% & 148 \\ 
\emph{Poa bulbosa} & 5 & 3\% & 1401 \\ 
\emph{Aegilops crassa} & 4 & 3\% & 170 \\ 
\emph{Atriplex prostrata} & 4 & 3\% & 112 \\ 
\emph{Brachypodium distachyon} & 4 & 3\% & 1581 \\ 
\emph{Fumaria densiflora} & 4 & 3\% & 194 \\ 
\emph{Gypsophila pilosa} & 4 & 3\% & 79 \\ 
\emph{Hordeum murinum} & 4 & 3\% & 1342 \\ 
\emph{Lathyrus aphaca} & 4 & 3\% & 2085 \\ 
\emph{Lathyrus sativus} & 4 & 3\% & 297 \\ 
\emph{Secale cereale} & 4 & 3\% & 1153 \\ 
\emph{Triticum durum} & 4 & 3\% & 219 \\ 
\emph{Vitis sylvestris} & 4 & 3\% & 3\textsuperscript{\textit{*}} \\ 
\emph{Adonis flammea} & 3 & 2\% & 79 \\ 
\emph{Avena sterilis} & 3 & 2\% & 50001\textsuperscript{\textit{*}} \\ 
\emph{Bassia arabica} & 3 & 2\% & 230 \\ 
\emph{Bromus sterilis} & 3 & 2\% & 385 \\ 
\emph{Cephalaria syriaca} & 3 & 2\% & 144 \\ 
\emph{Glaucium aleppicum} & 3 & 2\% & 22\textsuperscript{\textit{*}} \\ 
\emph{Helianthemum ledifolium} & 3 & 2\% & 668 \\ 
\emph{Lolium rigidum} & 3 & 2\% & 1464 \\ 
\emph{Phalaris paradoxa} & 3 & 2\% & 370 \\ 
\emph{Quercus ithaburensis} & 3 & 2\% & 1431 \\ 
\emph{Triticum aestivum compactum} & 3 & 2\% & 209 \\ 
\emph{Verbena officinalis} & 3 & 2\% & 1046 \\ 
\emph{Vicia narbonensis} & 3 & 2\% & 1529 \\ 
\emph{Triticum urartu} & — & — & 2033 \\ 
\emph{Aegilops speltoides} & — & — & 1429 \\ 
\emph{Aegilops tauschii} & — & — & 1395 \\ 
\bottomrule

\end{longtable}

\begin{minipage}{\linewidth}
\textsuperscript{\textit{*}}Excluded from modelling due to sample size\\
\end{minipage}
\endgroup

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia for the 56 modelled taxa (N=35208}

@@

\section{Model assessment}\label{model-assessment}

@@

@@

\caption{\label{fig-model-sensitivity-vs-n_occ}Model sensitivity by
number of training occurrences}

@@

Many taxa that occur (or are predicted to occur) across the `hilly
flanks' today---including most crop progenitors---are reconstructed to
have had a significantly more restricted distribution in the terminal
Pleistocene/Early Holocene. These include \emph{Ficus carica} (fig);
\emph{Hordeum} spp. (wild barleys); \emph{Lathyrus aphaca} and \emph{L.
sativus} (both marginally edible legumes); \emph{Triticum aestivum
compactum} (in the N. Levant), \emph{T. monococcum aegilopoides},
\emph{T. durum}, and \emph{Triticum urartu} (but not the other wheat
progenitor, \emph{T. turgidum dicoccum} -- see below); \emph{Aegilops
speltoides}, but not \emph{Aegilops tauschii} (goatgrasses); and
\emph{Vicia} spp. (vetches), including \emph{Vicia faba} (broad beans).
most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in
particular disappear from the predicted ranges of these species, leaving
the Levant and to a lesser extent the Aegean and Cyprus as refugia.

Our results for the Levant are consistent with the current understanding
of this region as developing early intensive foraging economies {[}the
Natufian culture, @{]} and as a centre of origin of agriculture {[}@{]}.
Within the Levant, retreat from the Badia.

Loss of the Northern Mesopotamia--Anatolia region from the predicted
ranges of crop progenitors is interesting in light of the `golden
triangle' hypothesis
\citep{Lev-YadunEtAl2000, KozlowskiAurenche2005, AbboEtAl2010}, which
puts this region at the centre of the development of agriculture.
Multiple lines of archaeological evidence have emerged that point away
from this hypothesis and towards a more geographically diverse origin
\citep{Asouti2006, cites}, but it remains the area with some of the
earliest clear evidence of domestication
\citep{ZoharyHopfWeiss2012, KabukcuEtAl2021, UlasEtAl2024}. Comparative
genetics also points to the Northern Mesopotamia region as the centre of
diversity of many crops \citep[e.g.][]{HaasEtAl2019}. But since these
studies are based on modern genomes, if the wild range of these plants
has, as our modelling suggests, shifted since the Early Holocene, the
apparent centre of diversity may have shifted with them. Our
reconstructions are consistent with the late arrival of intensive
plant-based foraging economies in this region (cf.~the Natufian of the
Levant), and more broadly there need not be a link between the core
\emph{wild} range of a plant and the core area of its domestication. A
scenario in which cultivation emerged at the edges of the ranges of
valuable plant resources---as a means of extend their natural niche---is
also plausible.

@@

Cyprus and the Aegean are not conventionally considered part of the
primary zone of domestication but rather amongst the first regions that
acquired agriculture from West Asia. Our analysis complicates this
picture, as it indicates that the wild ranges of many crop progenitors
included these regions. Early examples of several domesticates are
recorded at sites on Cyprus, Western Anatolia and Greece
\citep{ArranzOtaeguiRoe2023}, and the Aegean region was probably
connected to West Asia by a land bridge via Anatolia until the Early
Holocene \citep{AksuHiscott2022}. Were these area part of the same
broader `interaction sphere' \citep{cite} that produce Neolithic
agriculture in West Asia?

Exceptions to the dominant trend of range reduction include \emph{Cicer
reticulatum} (wild chickpea), which has a relatively stable range
centered on Northern Mesopotamia; and \emph{Triticum turgidum dicoccum}
(wild emmer wheat), which is predicted to have two limited ranges
centered around the Black Sea Coast of Anatolia and the Palmyra basin.
In the latter case, neither of these areas are part of the predicted
modern distribution of wild emmer (centered around the Caucasus and
Northern Mesopotamia), but it would be consistent with archaeological
evidence for early cultivation at sites in the Upper Euphrates
\citep{Willcox2024}.

@@

\begin{itemize}
\tightlist
\item
  Most cereal and legume crops predicted to be Levantine
\item
  Cereals show diverging southern/northern ranges (legumes not so much)
\item
  Not rye (Anatolia), chickpea (N. Mesopotamia)
\end{itemize}

Wild barley \textbf{?@fig-hindcast-barely-levant} and pistachio
\citep[\emph{Pistacia atlantica},][]{fig} show a contraction of their
predicted ranges from the Pleistocene to the Holocene, concurrent with
them being brought into cultivation. Plausible that this prompted
cultivation? These two species \citep{ArranzOtaeguiRoe2023} also see
marked declines in the archaeobotanical record from the Early PPNA/Early
PPNB (where they were amongst the most common taxa) to the Late PPNB and
Late Neolithic. However, this trend is not seen in most other crop
progenitors.

Cereals / other progenitors (not all) reduced in range before
cultivation? Reduction in range from BA to EH very evident in Hordeum,
less so in \ldots{}

@@

Is inconsistent with more ``macro'' trends such as the reduced range of
Hordeum and Pistacia correlating with its reduced abundance in the
archaeobotanical record.
